---
title: "DeepC Tutorial Format HiC Data"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, include=FALSE}

setwd("~/data_local/deepC_write_tutorials_offline/run_deepc_tutorial/")

```

This tutorial demonstrates how to format HiC data for deepC training. 
The tutorial walks through formating a single chromosome. In practice we prepare each 
chromosome separately and concatinate all chromosome files in the end to a single file for training.

### Global Set up

Load packages and source custom functions for deepC and general HiC processing and visualization.

```{r packages, message=FALSE}
library(tidyverse)
library(cowplot)
theme_set(theme_cowplot())
library(RColorBrewer)

source("./deepC/helper_for_preprocessing_and_analysis/functions_for_HiC.R")
source("./deepC/helper_for_preprocessing_and_analysis/functions_for_deepC.R")

```

Set global options, specifiying bin size and bp context of the model. 

```{r}
sample <- "GM12878"
bin.size <- 5000

window.size <- 1000000 + bin.size  # bp_context
# how many HiC skeleton bins have been used (10 for all published models)
interaction.value.bins <- 10  
# number of predictions bins (also called number of classes later). Refers to number 
# of bins in the vertial interaction pole
# depends on the bin.size and bp_context selected: 
# 201 for 5kb models, 101 for 10kb models 
prediction.bins <- window.size/bin.size 

# link to chromosome sizes file (ucsc style)
genome.sizes <- read.table('hg19_chrom_sizes.txt')  

```

```{bash}
head hg19_chrom_sizes.txt
```


### Read in HiC Data

Set the chromosome and read in the Hi-C data. The skeleton transformation is done chromosome wise and for memory friendliness we recommend doing this separate for each chromosome and concatinating the output into a single file.

We support two Hi-C data formats, both are sparse matrix formats: 

1) HiC-Pro style format comprising a sparse matrix with tab separated: *bin_id1 bin_id2 interaction_value* and a second bed-like file that lists the genomic coordinates of the bin_ids *chr start end bin_id*

2) A sparse contact matrix with the genomic coordinates as identifiers e.g.

```{bash}
head gm12878_primary_chr17_5kb.contacts.KRnorm.matrix
```
This matrix has norm factos applied thats why the contact values are not discrete.

Both formats list the left most base of a bin sized interval as identifier for the bin position.

We first read in the Hi-C data, inspect the object and create a binned genome template for filling.

```{r}

# Set chromosome name
chrom <- "chr17"

# get the respective chromosome size 
chrom.size <- as.numeric(genome.sizes[genome.sizes[,1] == chrom,2])

# to read in a HiC matrix with the format coord_bin1 coord_bin2 interaction_value 
# (e.g. as Rao et al. downloads) 
# provide the matrix file, the chromosome and bin.size arguments
# if the format is like HiC-Pro (bin1_id bin2_id interaction_value) with a separate 
# bed-like coordinate file
# provide the coordinate file in the "coords=" argument of the Import function
matrix.norm.loc <- "gm12878_primary_chr17_5kb.contacts.KRnorm.matrix"

hic <- ImportHicproMatrix(matrix.norm.loc, chr=chrom, bin.size=bin.size)

# remove all interactions that are more then bp_context apart
hic <- trimHicRange(hic, range = window.size + bin.size)

# The HiC data is stored in a structures list storing the bin size start position 
# and stat id as well as the coordinates with bin associations
str(hic)

#  and the actual contact matrix under $marix.df
hic$matrix.df

# check the first covered position in the HiC data
start.pos <- hic$start.pos - bin.size/2

# to map the hic data into consistent bins for deepC training, prediction and 
# plotting we create get a binned genome template this requires bedtools 
# to be installed as the R function calls bedtools in the background in bash
binned.genome <- getBinnedChrom(chr=chrom, start=start.pos, end=chrom.size, window=window.size, step=bin.size)
binned.genome <- as_tibble(binned.genome)
names(binned.genome) <- c("chr", "start", "end")
# filter out ibcombplete megabase bins
binned.genome <- binned.genome %>%
  mutate(width = end - start) %>%
  filter(width == window.size) %>%
  select(-width)

# quick look at the binned genome
binned.genome
  
```

### Convert to deepC Format

Next we convert the Hi-C interactions first into deepC format with a vvertical zig-zag pole associated with the center of each bp_context sized sequence window.
Second we transform the interaction values into the Hi-C skeleton by applying the pyramid shaped percentile binning. The data are percentile normalized into unequal bins from low to high values in the following scheme: 2x20% 4x10% 4x5% percentiles.

```{r}
# Convert Hi-C interaction matrix into genomic windows of bp_context size with the vertical 
# "zig-zag" pole of associated chromatin interactions
# Get ZigZiag Pole ZickZack Pole of interactions per window
#    /
#    \ 
#    /
#    \ 
#    |
# """""""
#     | 4 |     | 4 |
#       | 2 || 2 |
# | 5 | 3 | 1 | 3 | 5 | 
# uses helper perl scripts from the repository
# There is a memory friendly version of the function which relies on two perl helper scripts
# but takes a tad longer
tdf <- getZigZagWindowInteractionsPerl(hic, 
         binned.genome, 
         window.size, 
         bin.size,
         query.pl= "./deepC/helper_for_preprocessing_and_analysis/match_query_table.pl")
# tdf <- getZigZagWindowInteractionsPerlMemoryFriendlyFullPerl(hic, 
  # binned.genome, 
  # window.size, 
  # bin.size,
  # prepare.pl = "./deepC/current_version/......prepare_query_table.pl", #todo
  # query.pl= "./deepC/helper_for_preprocessing_and_analysis/match_query_table.pl")

# Impute zero values in interaction matrix with Median of a 5x5 neighborhood
# padd with median of respective interaction distance for calculating 
# the median on edge cases

# first add median column
tdf$median.value <- apply(tdf, 1, function(x){
  m <- median(as.numeric(x[c(4:length(x))]))
  return(m)
})

# filter out genomic windows with median 0 interaction values
fdf <- tdf[tdf$median.value > 0,]

# impute zeros with 5x5 median value
idf <- medianImputeZerosDataFrame(fdf, k=5)
idf <- idf[,-ncol(idf)]  #remove padded

# convert to tibble
idf <- as_tibble(idf)

# pyramid percentile binning
bdf <- pyramidBin(idf)

# inspect 
bdf

```

### Example Skeleton Plot

Now we plot a chunk of the skeleton for checking.

```{r, example_plot, fig.height=3, fig.width=9}
  # Create example plot
  # Make a position column, gather and tranform to numeric interactions bins
  # take first 1000 entries
  pdf <- head(bdf, n=1000) %>%
    mutate(pos = start+((end-start)/2)) %>%
    gather(bin, value, -chr, -start, -end, -pos, factor_key = T) %>%  # gather
    mutate(bin = as.numeric(bin)) %>%
    mutate(pos = if_else(bin %% 2 == 0, pos - bin.size/2, pos))  # adjust positions for 
    # zig-zag layout
  
  # # Plot Coverage of interaction windows
  tpdf <- triangularize(pdf, bin = bin.size)
  
  p <- ggplot(tpdf, aes(y=bin, x=pos, fill=value, group = polyid)) + 
    geom_polygon() + 
    scale_fill_gradientn(colours = brewer.pal(9, 'YlOrRd'))
  
  plot(p)
```

### Save

Now we just need to store the data in txt format. 

```{r}
  # collate to single comma separated column
  bdf$class <- bdf %>%
    select(-chr, -start, -end) %>%
    unite(col = class, sep = ",") %>%
    pull()

  # and trim output
  bdf <- bdf %>% select(chr, start, end, class)
  
  # inspect
  bdf
  
  # and save in plain txt
  write.table(bdf, file = paste0("hic_skeleton_for_deepC_5kb_", chrom,".bed"),
              col.names = F, row.names = F, quote = F, sep = "\t")
  
```


This will store a file per chromosome, which we then concatinate into a single file for all chromosomes and supply this to the deepC training routines. Find links to data fully formatted for training under *formatted_data_links*.



